---
title: "Intro to Bayesian Statistics"
session: 1
subtitle: "Part 3 <br/> Hierarchical models"
author: "Andrew Ellis"
# institute: "Kognitive Psychologie, Wahrnehmung und Methodenlehre, Universität Bern"
institute: "Methodenkurs Neurowissenschaft im Computerlab"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["css/xaringan-themer.css", "css/slides-style.css"]
    nature:
      highlightStyle: github #solarized-light # github
      highlightLines: true
      ratio: 16:10
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r child = "setup.Rmd"}
```

```{r load-packages, include=FALSE, warning=FALSE}
library(tidyverse)
library(rmarkdown)
library(countdown)
```



## Hierarchical Models

<!-- library(lmerTest) -->
<!-- options(contrasts=c('contr.sum', 'contr.poly')) -->
<!-- anova(lmer(y ~ x + (1|subj),data=myDat),ddf="Kenward-Roger") -->

<!-- library(afex) -->
<!-- mixed(y ~ x + (1|subj), type=3,method="KR",data=myDat)  -->


How can repeated measurements be taken into account?

--



---

class: middle

.pull-left-narrow[
  .huge-blue-number[1]
]
.pull-right-wide[
  .larger[
  Estimating Means
  ]
]

---

## Another IQ Example

This time we have 3 people, and 3 measurements from each.

.footnote[
Lee, Michael D., and Eric-Jan Wagenmakers. Bayesian Cognitive Modeling: A Practical Course. Cambridge: Cambridge University Press, 2014. https://doi.org/10.1017/CBO9781139087759.
]


.panelset[
.panel[.panel-name[Create data]

```{r}
IQwide <- tribble(
  ~A, ~B, ~C,
  110, 105, 115,
  105, 112, 108,
  102, 113, 130
)
```
]
.panel[.panel-name[Pivot longer]

```{r}
IQdata <- IQwide %>% 
  pivot_longer(everything(), names_to = "Person", values_to = "IQ") %>% 
  mutate(Person = as_factor(Person)) %>% 
  arrange(Person)
```
]
.panel[.panel-name[Data]

```{r}
library(kableExtra)
IQdata %>% 
  kbl() %>% 
  scroll_box(width = "500px", height = "200px")
```

]
]

---

## Point Estimates




```{r}
se <- function(x) sd(x)/sqrt(length(x))

IQdata %>% 
  group_by(Person) %>% 
  summarise(mean = mean(IQ),
            sd = sd(IQ),
            se = se(IQ))
```

---

```{r}
IQdata %>% 
  ggplot(aes(Person, IQ)) +
  geom_point()
```


---


.discussion[

- How would you analyse these data? 

]

```{r echo=FALSE}
countdown(minutes = 3)
```

---


## Parameter estimation


Our goal is to estimate the 3 latent IQ, and the group mean. We assume that for each person, the measurements are normally distributed around that person's latent IQ. For simplicity, we assume a common standard deviation.

$$ IQ_{i} \sim \mathcal{N}(\mu_j, \sigma^2) $$


```{r eval=FALSE, include=FALSE}
library(equatiomatic)
library(lme4)
mod1 <- lmer(IQ ~ 1 + (1 | Person), IQdata)

# Give the results to extract_eq
eq <- extract_eq(mod1) 
```
 
We have 3 possibilies:

- __No pooling:__ We can pretend that all people are independent of each other.

- __Complete pooling:__ We can pretend that all observations are from the same person.

- __Partial pooling:__ We can take into account that we have 3 observations for each person. These 3 people are similar to each other, i.e. they are samples from a super-population of people. 



---

## Partial Pooling Model

The partial pooling model can be written as

$$
\begin{aligned}
  \operatorname{IQ}_{i}  &\sim N \left(\alpha_{j}, \sigma^2 \right) \\
    \alpha_{j}  &\sim N \left(\mu_{\alpha}, \sigma^2_{\alpha} \right)
    \text{, for Person j = 1,} \dots \text{,J}
\end{aligned}
$$ 

where $\mu_{\alpha}$ is the population level effect, and the $\alpha_{j}$ are the latent IQ scores for each Person.

---

## No pooling



```{r}
library(brms)
get_prior(IQ ~ 0 + Person,
          data = IQdata)
```


---

## No pooling

```{r message=FALSE, warning=FALSE}
m_no_pool <- brm(IQ ~ 0 + Person,
                 data = IQdata,
                 file = "models/m_no_pool")
```


---

## Complete pooling


```{r}
get_prior(IQ ~ 1,
          data = IQdata)
```


---

## Complete pooling

```{r message=FALSE, warning=FALSE}
m_comp_pool <-  brm(IQ ~ 1,
                 data = IQdata,
                 file = "models/m_comp_pool")
```


---

## Partial pooling


```{r}
get_prior(IQ ~ 1 + (1 | Person),
                 data = IQdata)
```


---

## Partial pooling


```{r echo=FALSE}
library(tidybayes)
priors <- get_prior(IQ ~ 1 + (1 | Person),
                 data = IQdata)

priors %>%
  parse_dist(prior) %>%
  ggplot(aes(y = class, dist = .dist, args = .args)) +
  stat_dist_halfeye() +
  labs(
    title = "Prior Verteilungen",
    subtitle = "im Partial Pooling Modell",
    x = NULL
  )
```


---

## Partial pooling

```{r message=FALSE, warning=FALSE}
m_part_pool <-  brm(IQ ~ 1 + (1 | Person),
                 data = IQdata,
                 file = "models/m_part_pool")
```




---


.panelset[
.panel[.panel-name[No Pooling]

```{r}
m_no_pool
```

]

.panel[.panel-name[Complete Pooling]
```{r}
m_comp_pool
```

]
.panel[.panel-name[Partial Pooling]

```{r}
m_part_pool
```
]
]

---

## Partial Pooling

```{r}
f <- fixef(m_part_pool, summary = FALSE)
r <- ranef(m_part_pool, summary = FALSE)
```



```{r}
library(tidybayes)

get_variables(m_part_pool)

person_effects <- m_part_pool %>%
  spread_draws(b_Intercept, r_Person[Person, Intercept]) %>%
  # add the grand mean to the person-specific deviations
  mutate(mu = b_Intercept + r_Person)
```


---

```{r}
person_effects %>% 
  median_qi(mu)
```

```{r}
fixef(m_no_pool)
```

---

.panelset[
.panel[.panel-name[Plot Code]

```{r, echo=TRUE, eval=FALSE}
person_effects %>% 
  ggplot(aes(mu, Person)) +
  stat_halfeye(fill = "#A2BF8A") +
  geom_vline(xintercept = fixef(m_part_pool)[1, 1], color = "#839496", size = 1) +
  geom_vline(xintercept = fixef(m_part_pool)[1, 3:4], color = "#839496", linetype = 2) +
  labs(x = expression("Subject specific means"),
       y = "Subjects") +
  theme(panel.grid   = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y  = element_text(hjust = 0)) 
```
]
.panel[.panel-name[Plot]

```{r echo=FALSE}
person_effects %>% 
  ggplot(aes(mu, Person)) +
  stat_halfeye(fill = "#A2BF8A") +
  geom_vline(xintercept = fixef(m_part_pool)[1, 1], color = "#839496", size = 1) +
  geom_vline(xintercept = fixef(m_part_pool)[1, 3:4], color = "#839496", linetype = 2) +
  labs(x = expression("Subject specific means"),
       y = "Subjects") +
  theme(panel.grid   = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y  = element_text(hjust = 0)) 
```
]
]



---

## Partial Pooling vs No Pooling


Due to the hierarchical nature of the model, individual estimate are pulled toward the group mean. This means that the will tend to be less extreme. This is known as "shrinkage".


.panelset[
.panel[.panel-name[Plot Code]

```{r, echo=TRUE, eval=FALSE}
col <- viridis::viridis(3, begin = 0.2, end = 0.8)

person_effects %>% 
  ggplot(aes(mu, Person, fill = Person)) +
  stat_halfeye(alpha = 0.6) +
  geom_vline(xintercept = fixef(m_part_pool)[1, 1], color = "#ECCC87", 
             size = 1) +
  geom_vline(xintercept = fixef(m_no_pool)[1, 1], color = col[1], 
             size = 1) +
  geom_vline(xintercept = fixef(m_no_pool)[2, 1], color = col[2], 
             size = 1) +
   geom_vline(xintercept = fixef(m_no_pool)[3, 1], color = col[3], 
              size = 1) +
  scale_fill_viridis_d(begin = 0.2, end = 0.8) +
  labs(x = expression("Subject specific means),
       y = "Subjects") +
  theme(panel.grid   = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y  = element_text(hjust = 0)) 
```
]
.panel[.panel-name[Plot]

```{r, echo=FALSE}
col <- viridis::viridis(3, begin = 0.2, end = 0.8)

person_effects %>% 
  ggplot(aes(mu, Person, fill = Person)) +
  stat_halfeye(alpha = 0.6) +
  geom_vline(xintercept = fixef(m_part_pool)[1, 1], color = "#ECCC87", 
             size = 1) +
  geom_vline(xintercept = fixef(m_no_pool)[1, 1], color = col[1], 
             size = 1) +
  geom_vline(xintercept = fixef(m_no_pool)[2, 1], color = col[2], 
             size = 1) +
   geom_vline(xintercept = fixef(m_no_pool)[3, 1], color = col[3], 
              size = 1) +
  scale_fill_viridis_d(begin = 0.2, end = 0.8) +
  labs(x = expression("Subject specific means"),
       y = "Subjects") +
  theme(panel.grid   = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y  = element_text(hjust = 0)) 
```
]
]





---

class: middle

.pull-left-narrow[
  .huge-blue-number[2]
]
.pull-right-wide[
  .larger[
  Mixed Models
  ]
]

---

.panelset[
.panel[.panel-name[Create dataset]


```{r}
library(tidyverse)

intervention <- rep(c('treat', 'control'), each = 5)
pre <- c(20, 10, 60, 20, 10, 50, 10, 40, 20, 10)
post <- c(70, 50, 90, 60, 50, 20, 10, 30, 50, 10)
```

```{r}
dwide <- tibble(id = factor(1:10), 
            intervention, pre, post) %>% 
  mutate(diff = post - pre,
         id = as_factor(id), 
         intervention =  factor(intervention, levels = c("control", "treat")))
```
]
.panel[.panel-name[Dataframe]
```{r}
dwide %>% 
  rmarkdown::paged_table()
```

]]

---

```{r}
d <- dwide %>% 
  select(-diff) %>% 
  pivot_longer(cols = pre:post, names_to = "time", values_to = "score") %>% 
  mutate(time = as_factor(time))

d %>% 
  rmarkdown::paged_table()
```


---


We have 10 subjects.

.panelset[
.panel[.panel-name[Subjects]


```{r}
d %>% 
  summarize(id = n_distinct(id))
```
]

.panel[.panel-name[Trials per subject]

with 2 observations per subject.


```{r}
d %>% 
  group_by(id, intervention) %>% 
  count() %>% 
  rmarkdown::paged_table()
```
]
]

---

.discussion[
How would you analyse these data?
]

```{r echo=FALSE}
countdown(minutes = 5)
```

---

## Summarize Data

.panelset[
.panel[.panel-name[Standard error function]

```{r}
se <- function(x) sd(x)/sqrt(length(x))
```

]
.panel[.panel-name[Summarize (long)]

```{r}
d %>% 
  group_by(intervention, time) %>% 
  summarise(mean = mean(score),
            sd = sd(score),
            se = se(score))
```

]

.panel[.panel-name[Summarize (wide)]

```{r}
dwide %>% 
  group_by(intervention) %>% 
  summarise(mean = mean(diff),
            sd = sd(diff),
            se = se(diff))
```
]

]

---

.panelset[
.panel[.panel-name[Plot Code]

```{r echo=TRUE, eval=FALSE}
d %>% 
  ggplot(aes(time, score, color = intervention)) +
  geom_line(aes(group = id), linetype = 1, size = 1) +
  geom_point(size = 4) +
  scale_color_viridis_d(end = 0.8) +
  theme_bw()
```

]

.panel[.panel-name[Plot]

```{r echo=FALSE}
d %>% 
  ggplot(aes(time, score, color = intervention)) +
  geom_line(aes(group = id), linetype = 1, size = 1) +
  geom_point(size = 4) +
  scale_color_viridis_d(end = 0.8) +
  theme_bw()
```

]
]



---

## Welch Test

```{r}
t.test(diff ~ intervention,
       data = dwide, 
       var.equal = FALSE)
```

---


## Mixed Model



.panelset[
.panel[.panel-name[Model Code]

```{r}
library(lme4)
lme_model <- lmer(score ~ intervention + time + (1|id), 
                  data = d)
```

]
.panel[.panel-name[Model summary]
```{r}
lme_model %>% 
  sjPlot::tab_model()
```
]
.panel[.panel-name[Model equations]

$$
\begin{aligned}
  \operatorname{score}_{i}  &\sim N \left(\alpha_{j[i]} + \beta_{1}(\operatorname{time}_{\operatorname{post}}), \sigma^2 \right) \\
    \alpha_{j}  &\sim N \left(\gamma_{0}^{\alpha} + \gamma_{1}^{\alpha}(\operatorname{intervention}_{\operatorname{treat}}) + \gamma_{2}^{\alpha}(\operatorname{intervention}_{\operatorname{treat}} \times \operatorname{time}_{\operatorname{post}}), \sigma^2_{\alpha_{j}} \right)
    \text{, for id j = 1,} \dots \text{,J}
\end{aligned}
$$
]

---


```{r}
library(afex)
mixed(score ~ intervention * time + (1|id), 
      type = 3, method = "KR",
      data = d)
```

---


## What is going on here?

```{r}
mm <- model.matrix(~ intervention * time,
             data = d)
head(mm)
```

---

## Erklärung

- We are predicting the mean of the outcome variable `score` with a linear model. 

- The mean is decomposed into 
 + population-level effects: `intervention * time`
 + group-level effects (a deviation from the average for every person): `(1 | id)`

---


## Using `brms`


.panelset[
.panel[.panel-name[get_prior]

```{r}
library(brms)

priors1 <- get_prior(score ~ intervention*time,
          data = d)

```
]
.panel[.panel-name[Priors]

```{r}
priors1 %>% 
    as_tibble() %>% 
    select(1:4)
```

]
]

---

```{r}
priors2 <- get_prior(score ~ intervention*time + (1 | id),
          data = d) 
```

```{r}
priors2 %>% 
    as_tibble() %>% 
    select(1:4)
```

---

```{r}
m2 <- brm(score ~ intervention*time + (1 | id),
          data = d,
          file =  "models/04-treat-time") 
```

---

```{r}
summary(m2)
```


---

```{r}
m2 %>% 
  mcmc_plot("b_")
```

                                                                                                                                                                                    
---

```{r}
conditional_effects(m2, effects = "intervention:time")
```



