---
title: "Multilevel models"
description: | 
  Parameter estimation in clustered data
date: 05-27-2021
author:
  - first_name: "Andrew"
    last_name: "Ellis"
    url: https://github.com/awellis
    affiliation: Kognitive Psychologie, Wahrnehmung und Methodenlehre, Universität Bern 
    affiliation_url: https://www.kog.psy.unibe.ch
    orcid_id: 0000-0002-2788-936X

citation_url: https://awellis.github.io/learnmultilevelmodels/multilevel-models.html
bibliography: ./bibliography.bib
output: 
    distill::distill_article:
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


:::note
I am using the native `pipe` operator, which is new to R 4.10. This pipe operator is written as a `|` followed by a `>`. In this document, the operator is printed as `|>`, due to the fact that I am using font ligatures. If the pipe doesn't work for you, simply replace it with the older pipe `%>%`.
:::

```{r}
library(tidyverse)
library(brms)
library(tidybayes)

theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))
```


```{r}
set.seed(5)

n_obs <- 10
n_conditions <-  3
condition_means <- c(0, 1, 2)
sigma <- 0.5

d <- tibble(condition = rep(LETTERS[1:n_conditions], n_obs),
            response = rnorm(n_obs * n_conditions, condition_means, sigma)) |> 
  arrange(condition)
```


```{r}
p1 <- d |> 
  ggplot(aes(response, condition, color = condition)) +
  geom_point( size = 2) +
  scale_color_brewer(type = "qual")

p1
```

```{r}
condition_means <- d |> 
  group_by(condition) |> 
  summarise(mean = mean(response),
            sd = sd(response))
```


```{r}
p1 +
  geom_point(aes(mean, condition, color = condition), 
             data = condition_means, 
             size = 6)
```



```{r}
formula1 <- response ~ 0 + condition
```


```{r eval=FALSE, include=FALSE}
library(equatiomatic)
lm1 <- lm(formula1, data = d)
eq1 <- extract_eq(lm1)
```

$$
\operatorname{response} = \beta_{\operatorname{A}}(\operatorname{condition}_{\operatorname{A}}) + \beta_{\operatorname{B}}(\operatorname{condition}_{\operatorname{B}}) + \beta_{\operatorname{C}}(\operatorname{condition}_{\operatorname{C}}) + \epsilon
$$


```{r}
get_prior(formula1, data = d) |> 
  as_tibble() |> select(1:4)
```


```{r}
priors1 <- prior(normal(1, 1), class = b)

tibble(x = seq(-3, 5, by = 0.01),
       y = dnorm(x, 1, 1)) |> 
  ggplot(aes(x, y)) + geom_line(size = 2) +
  geom_vline(xintercept = 1, linetype = 2) +
  ylab("") + xlab("") +
  ggtitle("Prior on means")

```



```{r}
m1 <- brm(formul, 
          prior = priors1,
          data = d,
          file = "models/02-m1")
```



```{r}
summary(m1)
```

```{r}
mcmc_plot(m1, type = "areas")
```

```{r}
fixef(m1)
```

```{r}
m1_pop <- fixef(m1, summary = FALSE)
```

```{r}
m1_pop |> 
  as_tibble() |> 
  pivot_longer(everything(),
               names_to = "condition",
               values_to = "mean") |> 
  group_by(condition) |> 
  summarise(mean = mean(mean))
```

```{r}
conditional_effects(m1)
```




## Hierarchical model

```{r}
condition_means

standard_deviation <- function(x) {
    sqrt(sum((x - mean(x))^2) / (length(x)))
}

condition_means <- d |> 
  group_by(condition) |> 
  summarise(mean = mean(response),
            sd = sd(response))

condition_means |> 
  summarize(mu = mean(mean),
            sd = sd(mean))
```


```{r}
formula2 <- response ~ 1 + (1 | condition)
```


```{r eval=FALSE, include=FALSE}
lm2 <- lme4::lmer(formula2, data = d)
eq2 <- extract_eq(lm2)
```


$$
\begin{aligned}
  \operatorname{response}_{i}  &\sim N \left(\alpha_{j[i]}, \sigma^2 \right) \\
    \alpha_{j}  &\sim N \left(\mu_{\alpha_{j}}, \sigma^2_{\alpha_{j}} \right)
    \text{, for condition j = 1,} \dots \text{,J}
\end{aligned}
$$


```{r}
get_prior(formula2, data = d) |> 
  as_tibble() |> select(1:4)
```

```{r}
tibble(x = seq(0, 15, by = 0.01),
       Intercept = dstudent_t(x, 3, 0.9, 2.5),
       sd = dstudent_t(x, 3, 0, 2.5)) |> 
  pivot_longer(c(Intercept, sd), names_to = "prior", values_to = "y") |> 
  ggplot(aes(x, y, color = prior)) + geom_line(size = 2) +
  ylab("") + xlab("") +
  scale_color_brewer(type = "qual") +
  ggtitle("Intercept")
```


```{r}
m2 <- brm(formula2,
          data = d,
          control = list(adapt_delta = 0.9),
          file = "models/02-m2")
```

```{r}
summary(m2)
```

```{r}
mcmc_plot(m2, type = "areas")
```

```{r}
priors3 <- prior(normal(1, 1), class = Intercept) +
  prior(student_t(3, 0, 2.5), class = sd)

m3 <- brm(formula2,
          data = d,
          prior = priors3,
          control = list(adapt_delta = 0.9),
          file = "models/02-m3")
```


```{r}
summary(m3)
```

```{r}
priors4 <- prior(normal(1, 1), class = Intercept) +
  prior(student_t(3, 0, 2.5), class = sd)


tibble(x = seq(0, 15, by = 0.01),
       weak = dstudent_t(x, 3, 0, 2.5),
       strong = dstudent_t(x, 3, 0, 1)) |> 
  pivot_longer(c(weak, strong), names_to = "prior", values_to = "y") |> 
  ggplot(aes(x, y, color = prior)) + geom_line(size = 2) +
  ylab("") + xlab("") +
  scale_color_brewer(type = "qual") +
  ggtitle("Intercept")
```


```{r}
m4 <- brm(formula2,
          data = d,
          prior = priors4,
          control = list(adapt_delta = 0.9),
          file = "models/02-m4")
```


```{r}
summary(m4)
```




```{r eval=FALSE, include=FALSE}
posts <-
  tibble(name  = str_c("fit", 1:3),
         prior = str_c("normal(0, ", c(10, 1, 0.1), ")")) |> 
  mutate(fit = map(name, get)) |> 
  mutate(post = map(fit, posterior_samples))
  
head(posts)
```

```{r}
posts <-
  tibble(name  = str_c("m", 2:4),
         model = str_c("Model ", 2:4)) |> 
  mutate(fit = map(name, get)) |> 
  mutate(post = map(fit, posterior_samples))
  
# head(posts)

posts <- posts |> 
  select(-fit) |> 
  unnest(post)

posts_mean <- posts |> 
  pivot_longer(starts_with("b_Intercept"), names_to = "condition",
               values_to = "value")
posts_mean |> 
  ggplot(aes(x = value, y = model)) +
  stat_halfeye() +
  facet_wrap(~ condition)
```

```{r}
posts_sd <- posts |> 
  pivot_longer(starts_with("sd_condition__Intercept"), names_to = "condition",
               values_to = "value")
posts_sd |> 
  ggplot(aes(x = value, y = model)) +
  stat_halfeye() +
  facet_wrap(~ condition)
```


### varying effects

```{r}
m4 |> 
  spread_draws(r_condition[condition])
```

```{r}
m4 %>%
  spread_draws(b_Intercept, r_condition[condition]) |> 
  median_qi(condition_mean = b_Intercept + r_condition) 
```

```{r}
p4_shrinkage <- m4 |> 
  spread_draws(b_Intercept, r_condition[condition]) |> 
  median_qi(condition_mean = b_Intercept + r_condition) |> 
  ggplot(aes(y = condition, x = condition_mean, 
             color = condition)) +
  geom_pointinterval(aes(xmin = .lower, xmax = .upper), point_size = 4) +
  scale_color_brewer(type = "qual")
p4_shrinkage 
```

```{r}

p4_shrinkage +
    geom_point(aes(mean, condition), 
             data = condition_means, 
             size = 2, color = "black") +
  ggtitle("Hierarchical shrinkage")
```




## Multilevel example


```{r}
simulate_treament <- function(a = 3.5,
                          b = -1,
                          sigma_a = 1,
                          sigma_b = 0.8,
                          rho = -0.7,
                          n_subjects = 10,
                          n_trials = 20,
                          sigma = 0.5) {

    # Ccombine the terms
    mu <- c(a, b)
    cov_ab <- sigma_a * sigma_b * rho
    SD  <- matrix(c(sigma_a^2, cov_ab,
                       cov_ab, sigma_b^2), ncol = 2)

   #  sigmas <- c(sigma_a, sigma_b)          # standard deviations
   #  rho <- matrix(c(1, rho,             # correlation matrix
   #                 rho, 1), nrow = 2)
   # 
   # # now matrix multiply to get covariance matrix
   # SD <- diag(sigmas) %*% rho %*% diag(sigmas)

    varying_effects <-
        MASS::mvrnorm(n_subjects, mu, SD) |>
        # as_tibble(.name_repair = "unique") |>
        data.frame() |>
        purrr::set_names("a_j", "b_j")

    d_linpred <-
        varying_effects |>
        mutate(subject  = 1:n_subjects) |>
        expand(nesting(subject, a_j, b_j), post = c(0, 1)) |>
        mutate(mu = a_j + b_j * post,
               sigma = sigma) |>
        mutate(treatment = ifelse(post == 0, "pre", "post"),
               treatment = factor(treatment, levels = c("pre", "post")))

    d <- d_linpred |>
        slice(rep(1:n(), each = n_trials)) |>
        mutate(response = rnorm(n = n(), mean = mu, sd = sigma))

   d
}
```

```{r}
plot_linpred <- function(d, violin = FALSE) {
  
  d_linpred <- d |>
  group_by(subject, treatment) |> distinct(mu, .keep_all = TRUE)
  
  p <- d_linpred |>
    ggplot(aes(x = treatment, y = mu))
  
  if (isTRUE(violin)) {
  p <- p + 
      geom_violin(aes(x = treatment, y = response,
                          fill = treatment), 
                  alpha = 0.5,
                  data = d) +
      geom_jitter(aes(x = treatment, y = response,
                          color = treatment), 
                  width = 0.1, size = 2,
                  data = d)
  }
  
  p <- p +
    geom_line(aes(group = 1), color = "#8B9DAF", size = 1, linetype = 3) +
    geom_point(aes(fill = treatment), 
               shape = 21, 
               colour = "black", 
               size = 4, 
               stroke = 2) +
    scale_fill_brewer(type = "qual") +
    scale_color_brewer(type = "qual") +
    coord_cartesian(ylim = c(0, 8)) +
    ylab("Response") +
    theme(legend.position = "none",
          axis.ticks.x    = element_blank()) +
    facet_wrap(~ subject) +
  ggtitle("Linear predictor")
  
  p
}

```


```{r}
set.seed(52)
d1 <- simulate_treament(n_subjects = 5)
```

```{r}
plot_linpred(d1)
```


```{r}
plot_linpred(d1, violin = T)
```

```{r}
set.seed(52)
d2 <- simulate_treament(n_subjects = 5, 
                        sigma_a = 1.2, 
                        sigma_b = 0.2, 
                        rho = -0.7)

plot_linpred(d2, violin = F)
```


```{r}
d <- simulate_treament(n_subjects = 10, n_trials = 50)
plot_linpred(d, violin = F)
```


\begin{align*}
\text{wait}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i         & = \alpha_{\text{cafe}_i} + \beta_{\text{cafe}_i} \text{afternoon}_i \\
\begin{bmatrix} \alpha_\text{cafe} \\ \beta_\text{cafe} \end{bmatrix} & \sim \text{MVNormal} \left (\begin{bmatrix} \alpha \\ \beta \end{bmatrix}, \mathbf{S}  \right ) \\
\mathbf S     & = \begin{bmatrix} \sigma_\alpha & 0 \\ 0 & \sigma_\beta \end{bmatrix} \mathbf R \begin{bmatrix} \sigma_\alpha & 0 \\ 0 & \sigma_\beta \end{bmatrix} \\
\alpha        & \sim \operatorname{Normal}(0, 10) \\
\beta         & \sim \operatorname{Normal}(0, 10) \\
\sigma        & \sim \operatorname{HalfCauchy}(0, 1) \\
\sigma_\alpha & \sim \operatorname{HalfCauchy}(0, 1) \\
\sigma_\beta  & \sim \operatorname{HalfCauchy}(0, 1) \\
\mathbf R     & \sim \operatorname{LKJcorr}(2),
\end{align*}


```{r}
get_prior(response ~ treatment + (treatment | subject),
          data = d) |> 
  as_tibble() |> select(1:4)

```


```{r}
library(rethinking)

n_sim <- 1e5

set.seed(13)
r_1 <- 
  rlkjcorr(n_sim, K = 2, eta = 1) %>%
  as_tibble()

set.seed(13)
r_2 <- 
  rlkjcorr(n_sim, K = 2, eta = 2) %>%
  as_tibble()

set.seed(13)
r_4 <- 
  rlkjcorr(n_sim, K = 2, eta = 4) %>%
  as_tibble()
```

```{r}
ggplot(data = r_1, aes(x = V2)) +
  geom_density(color = "transparent", fill = "#5e81ac", alpha = 2/3) +
  geom_density(data = r_2,
               color = "transparent", fill = "#a3be8c", alpha = 2/3) +
  geom_density(data = r_4,
               color = "transparent", fill = "#bf616a", alpha = 2/3) +
  geom_text(data = tibble(x = c(.83, .62, .46),
                          y = c(.54, .74, 1),
                          label = c("eta = 1", "eta = 2", "eta = 4")),
            aes(x = x, y = y, label = label),
            color = "#A65141", family = "Courier") +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("correlation")
```


```{r}

fit1_prior <- brm(response ~ treatment + (treatment | subject),
            prior = prior(normal(0, 10), class = b),
            data = d,
            sample_prior = "only",
            file = "models/02-fit1-prior")

fit1 <- brm(response ~ treatment + (treatment | subject),
            prior = prior(normal(0, 10), class = b),
            data = d,
            file = "models/02-fit1")
```

```{r}
summary(fit1)
```


```{r}
mcmc_plot(fit1_prior)
 ```

```{r}
mcmc_plot(fit1)
```
```{r}
pp_check(fit1_prior, type = "dens_overlay_grouped", group = "treatment")
```

```{r}
pp_check(fit1, type = "dens_overlay_grouped", group = "treatment")
```


```{r}
grid <- d |> 
  modelr::data_grid(subject, treatment)

grid

fits <- grid %>%
  add_fitted_draws(fit1)

preds <- grid %>%
  add_predicted_draws(fit1)

d %>%
  ggplot(aes(y = treatment, x = response)) +
  stat_interval(aes(x = .prediction), data = preds) +
  stat_pointinterval(aes(x = .value), data = fits, .width = c(.66, .95), 
                     position = position _nudge(y = -0.3)) +
  geom_point() +
  scale_color_brewer()
```
