---
title: "Assignment 4"
description: | 
  Effect of tDCS stimulation on episodic memory
date: 05-28-2021
author:
  - first_name: "Andrew"
    last_name: "Ellis"
    url: https://github.com/awellis
    affiliation: Kognitive Psychologie, Wahrnehmung und Methodenlehre, Universität Bern 
    affiliation_url: https://www.kog.psy.unibe.ch
    orcid_id: 0000-0002-2788-936X

citation_url: https://awellis.github.io/learnmultilevelmodels/asssignment-4.html
bibliography: ./bibliography.bib
output: 
    distill::distill_article:
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: false
      self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
library(tidyverse)
library(brms)

theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))
```




# Effect of tDCS on memory


You are going to analyze data from a study in which 34 elderly subjects were given anodal (activating) tDCS over their left tempero-parietal junction (TPJ). Subjects were instructed to learn association between images and pseudo-words (data were inspired by @antonenkoTDCSinducedEpisodicMemory2019).

Episodic memory is measured by percentage of correctly recalled word-image pairings. We also have response times for correct decisions.

Each subject was tested 5 times in the TPJ stimulation condition, and a further 5 times in a sham stimulation condition.

The variables in the dataset are:

```
subject: subject ID
stimulation: TPJ or sham (control)
block: block
age: age
correct: accuracy per block
rt: mean RTs for correct responses
```

You are mainly interested in whether recall ability is better during the TPJ stimulation condition than during sham.


```{r}
d <- read_csv("https://raw.githubusercontent.com/kogpsy/neuroscicomplab/main/data/tdcs-tpj.csv")
```


```{r}
glimpse(d)
```

```{r}
d <- d %>% 
  mutate(across(c(subject, stimulation, block), ~as_factor(.))) %>% 
  drop_na()
```



:::exercise

- specify a linear model

- specify a null model

- estimate a Bayes factor for the effect of TPJ stimulation
    + using the Savage-Dickey density ratio test
    + using the bridge sampling method

- control for subjects' age
:::

