---
title: "Assignment 2"
description: | 
  Parameter estimation in clustered data
date: 05-28-2021
author:
  - first_name: "Andrew"
    last_name: "Ellis"
    url: https://github.com/awellis
    affiliation: Kognitive Psychologie, Wahrnehmung und Methodenlehre, Universität Bern 
    affiliation_url: https://www.kog.psy.unibe.ch
    orcid_id: 0000-0002-2788-936X

citation_url: https://awellis.github.io/learnmultilevelmodels/asssignment-2.html
bibliography: ./bibliography.bib
output: 
    distill::distill_article:
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: false
      self_contained: false
---


```{r}
library(tidyverse)
library(brms)

theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))
```



## Exercise 1

Use the `simulate_treament()` function to generate data. Vary the parameter settings, and then attempt to recover the known parameters using `brm()`. 

## Exercise 2

Use the `simulate_treament()` function to generate data. Give your data to a colleague and  they will give you theirs in return. Try to recover the (unknown to you) parameters using `brm()`. 



## Exercise 3
 

Load the dataset `sleepstudy` from the `lme4` package. These are data from a  sleep deprivation study, in which the average reaction time per day (in milliseconds) was recorded.

Have a look at the data, and then try to estimate the effect of `day` on reaction time.

:::note
It's always a good idea to plot the data before you do anything else.
:::

```{r}
library(lme4)
glimpse(sleepstudy)
```



```{r eval=FALSE, include=FALSE}
fit_sleepstudy <- brm(Reaction ~ Days + (1 + Days|Subject), 
                       data = sleepstudy)
summary(fit_sleepstudy)
```

