---
title: "Longitudinal Models"
description: | 
  Modelling change over time
date: 06-11-2021
author:
  - first_name: "Andrew"
    last_name: "Ellis"
    url: https://github.com/awellis
    affiliation: Cognitive psychology, perception & methods, Univerity of Bern 
    affiliation_url: https://www.kog.psy.unibe.ch
    orcid_id: 0000-0002-2788-936X

citation_url: https://awellis.github.io/learnmultilevelmodels/longitudinal-models.html
bibliography: ./bibliography.bib
output: 
    distill::distill_article:
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: false
      css: ./css/style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


The examples in this document are taken from @singerAppliedLongitudinalData and the translation into tidyverse and brms code by @kurzAppliedLongitudinalDataAnalysis2021.


We will first briefly look at how to reshape datasets, and then investigate two examples of longitudinal data with a hierarchical structure.


```{r packages}
library(kableExtra)
library(tidyverse)
library(brms)

# set ggplot theme
theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))

# set rstan options
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = 4)
```





# Exlporing data


We'll load a dataset from @raudenbushGrowthCurveAnalysis1992.

```{r}
tolerance <- read_csv("https://stats.idre.ucla.edu/wp-content/uploads/2016/02/tolerance1.txt", col_names = TRUE)

head(tolerance, n = 16)
```
The data are in the wide format (one row per `id`), but we need one row per observation, so we need to rehape the data into a long format.



```{r}
tolerance <- tolerance |> 
  pivot_longer(-c(id, male, exposure),
               names_to = "age", 
               values_to = "tolerance") |> 
  # remove the `tol` prefix from the `age` values 
  mutate(age = str_remove(age, "tol") |> as.integer()) |> 
  arrange(id, age) 
```



We have 16 subjects.

```{r}
tolerance %>% 
  distinct(id) %>% 
  count()
```

```{r}
tolerance %>%
  slice(c(1:9, 76:80))
```



First, we will plot the observations over time (at different ages), and connect them with lines.


```{r}
tolerance %>%
  ggplot(aes(x = age, y = tolerance)) +
  geom_point() +
  geom_line() +
  coord_cartesian(ylim = c(1, 4)) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~id)
```



Next, we will apply locally weighted smoothing.


```{r}
tolerance %>%
  ggplot(aes(x = age, y = tolerance)) +
  geom_point() +
  stat_smooth(method = "loess", se = F, span = .9) +
  coord_cartesian(ylim = c(1, 4)) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~id)
```



And finally, we'll estimate the linear effect of age.

```{r}
tolerance %>%
  ggplot(aes(x = age, y = tolerance)) +
  geom_point() +
  stat_smooth(method = "lm", se = F, span = .9) +
  coord_cartesian(ylim = c(1, 4)) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~id)
```




# Early Intervention

We'll start with a data set on early intervention on child development [@singerAppliedLongitudinalData]


> As part of a larger study of the effects of early intervention on child development, these researchers tracked the cognitive performance of 103 African- American infants born into low-income families. When the children were 6 months old, approximately half (n = 58) were randomly assigned to participate in an intensive early intervention program designed to enhance their cognitive functioning; the other half (n = 45) received no intervention and constituted a control group. Each child was assessed 12 times between ages 6 and 96 months. Here, we examine the effects of program participation on changes in cognitive performance as measured by a nationally normed test administered three times, at ages 12, 18, and 24 months.

> Each child has three records, one per wave of data collection. Each record contains four variables: (1) ID; (2) AGE, the child’s age (in years) at each assessment (1.0, 1.5, or 2.0); (3) COG, the child’s cognitive performance score at that age; and (4) PROGRAM, a dichotomy that describes whether the child participated in the early intervention program. Because children remained in their group for the duration of data collection, this predictor is time-invariant.
 
 
```{r}
early_int <- read_csv("data/early-intervention.csv") |> 
  mutate(id = as_factor(id),
         intervention = factor(ifelse(program == 0, "no", "yes"),
                               levels = c("no", "yes")))

head(early_int, 10)
```



In addition, we have the `age-1` variable, `age_c`. This variable has the value 0 when the child is 1 year old.

```{r}
early_int |> 
  filter(id %in% sample(levels(early_int$id), 12)) |> 
  ggplot(aes(x = age, y = cog)) +
  stat_smooth(method = "lm", se = F) +
  geom_point() +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~id, ncol = 4)
```


Our model for each child's development is:

$$
\text{cog}_{ij} = [ \pi_{0i} + \pi_{1i} (\text{age}_{ij} - 1) ] + [\epsilon_{ij}].
$$


\begin{align*}
\text{cog} & \sim \operatorname{Normal} (\mu_{ij}, \sigma_\epsilon^2) \\
\mu_{ij}   & = \pi_{0i} + \pi_{1i} (\text{age}_{ij} - 1).
\end{align*}




The $i^{th}$ child's `cog` score at observation $j$ is normally distributed, with mean $\mu$ and residuak standard deviation $\sigma_{\epsilon}$. The mean $\mu$ is modelled as a linear function of age. The intercept will correspond to the expected score at age 1.



Next, we consider that the children are assigned to two different group.

```{r}
early_int<-
  early_int|> 
  mutate(label = str_c("Intervetion = ", intervention)) 

early_int |> 
  ggplot(aes(x = age, y = cog, color = label)) +
  stat_smooth(aes(group = id),
              method = "lm", se = F, size = 1/6) +
  stat_smooth(method = "lm", se = F, size = 2) +
  scale_color_viridis_d(option = "B", begin = .33, end = .67) +
  scale_x_continuous(breaks = c(1, 1.5, 2)) +
  ylim(50, 150) +
  theme(legend.position = "none",
        panel.grid = element_blank()) +
  facet_wrap(~label)
```



## Fitting the model with brms

```{r}
priors_early_1 <- prior(normal(0, 20), class = b)

fit_early_1 <-
  brm(cog ~ 0 + Intercept + age_c + intervention + age_c:intervention + 
        (1 + age_c | id),
      prior = priors_early_1,
      data = early_int,
      control = list(adapt_delta = 0.95),
      seed = 3,
      file = "models/fit_early_1",
      file_refit = "on_change")
```





```{r}
fit_early_1
```

```{r}
conditional_effects(fit_early_1,
                    "age_c:intervention")
```



:::exercise

How would you go about demonstrating evidence for or against a difference between the interventions?
:::




# Changes in adolescent alcohol use


> As part of a larger study of substance abuse, Curran, Stice, and Chassin (1997) collected three waves of longitudinal data on 82 adolescents. Each year, beginning at age 14, the teenagers completed a four-item instrument assessing their alcohol consumption during the previous year. Using an 8-point scale (ranging from 0 = "not at all" to 7 = "every day"), adolescents described the frequency with which they (1) drank beer or wine, (2) drank hard liquor, (3) had five or more drinks in a row, and (4) got drunk. The data set also includes two potential predictors of alcohol use: COA, a dichotomy indicating whether the adolescent is a child of an alcoholic parent; and PEER, a measure of alcohol use among the adolescent’s peers. This latter predictor was based on information gathered during the initial wave of data collection. Participants used a 6-point scale (ranging from 0 = "none" to 5 = "all") to estimate the proportion of their friends who drank alcohol occasionally (one item) or regularly (a second item).

> Do individual trajectories of alcohol use during adolescence differ according to the history of parental alcoholism and early peer alcohol use.


```{r}
library(tidyverse)

alcohol <- read_csv("data/alcohol1_pp.csv")
head(alcohol)
```

```{r echo=FALSE}
tribble(~Variable, ~Description,
  "id", "subject ID",
  "age", "in years",
  "coa", "child of alcoholic parent (1 - yes, 0 - no)",
  "male", "indicator for male",
  "age_14", "age - 14 (0 corresponds to age: 14)",
  "alcuse", "alcohol consumption",
  "peer", "alcohol use among peers",
  "cpeer, ccoa", "centred variables") |> 
  kbl() |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```






```{r}
alcohol %>%
  filter(id %in% c(4, 14, 23, 32, 41, 56, 65, 82)) %>%
  
  ggplot(aes(x = age, y = alcuse)) +
  stat_smooth(method = "lm", se = F) +
  geom_point() +
  coord_cartesian(xlim = c(13, 17),
                  ylim = c(-1, 4)) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~id, ncol = 4)
```



\begin{align*}
\text{alcuse}_{ij} & = \pi_{0i} + \pi_{1i} (\text{age}_{ij} - 14) + \epsilon_{ij}\\
\epsilon_{ij}      & \sim \operatorname{Normal}(0, \sigma_\epsilon^2),
\end{align*}


\begin{align*}
\text{alcuse}_{ij} & = \big [ \gamma_{00} + \gamma_{10} \text{age_14}_{ij} + \gamma_{01} \text{coa}_i + \gamma_{11} (\text{coa}_i \times \text{age_14}_{ij}) \big ] \\
& \;\;\;\;\; + [ \zeta_{0i} + \zeta_{1i} \text{age_14}_{ij} + \epsilon_{ij} ] \\
\epsilon_{ij} & \sim \operatorname{Normal} (0, \sigma_\epsilon^2) \\

\begin{bmatrix} \zeta_{0i} \\ \zeta_{1i} \end{bmatrix} & \sim \operatorname{Normal} 
\begin{pmatrix}
\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
\begin{bmatrix} \sigma_0^2 & \sigma_{01} \\ \sigma_{01} & \sigma_1^2 \end{bmatrix}
\end{pmatrix},
\end{align*}




## Unconditional means model

\begin{align*}
\text{alcuse}_{ij} & =  \gamma_{00} +  \zeta_{0i} + \epsilon_{ij} \\
\epsilon_{ij}      & \sim \operatorname{Normal}(0, \sigma_\epsilon^2) \\
\zeta_{0i}         & \sim \operatorname{Normal}(0, \sigma_0^2).
\end{align*}


```{r}
get_prior(alcuse ~ 1 + (1 | id),
          data = alcohol)
```

```{r}
fit_alcuse_1 <-
  brm(alcuse ~ 1 + (1 | id),
      data = alcohol,
      file = "models/fit_alcuse_1",
      file_refit = "on_change")
```


## Unconditional growth model

\begin{align*}
\text{alcuse}_{ij} & = \gamma_{00} + \gamma_{10} \text{age_14}_{ij} + \zeta_{0i} + \zeta_{1i} \text{age_14}_{ij} + \epsilon_{ij} \\
\epsilon_{ij} & \sim \operatorname{Normal} (0, \sigma_\epsilon^2) \\
\begin{bmatrix} \zeta_{0i} \\ \zeta_{1i} \end{bmatrix} & \sim \operatorname{Normal} 
\begin{pmatrix}
\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
\begin{bmatrix} \sigma_0^2 & \sigma_{01} \\ \sigma_{01} & \sigma_1^2 \end{bmatrix}
\end{pmatrix}.
\end{align*}



```{r}
get_prior(alcuse ~ 0 + Intercept + age_14 + (1 + age_14 | id),
          data = alcohol)
```


```{r fit4.2}
fit_alcuse_2 <-
  brm(alcuse ~ 0 + Intercept + age_14 + (1 + age_14 | id),
      prior = c(prior(normal(0, 15), class = b)),
      data = alcohol,
      file = "models/fit_alcuse_2")
```



## Effect of COA

\begin{align*}
\text{alcuse}_{ij} & = \gamma_{00} + \gamma_{01} \text{coa}_i + \gamma_{10} \text{age_14}_{ij} + \gamma_{11} \text{coa}_i \times \text{age_14}_{ij} + \zeta_{0i} + \zeta_{1i} \text{age_14}_{ij} + \epsilon_{ij} \\
\epsilon_{ij} & \sim \text{Normal} (0, \sigma_\epsilon^2) \\
\begin{bmatrix} \zeta_{0i} \\ \zeta_{1i} \end{bmatrix} & \sim \text{Normal} 
\begin{pmatrix}
\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
\begin{bmatrix} \sigma_0^2 & \sigma_{01} \\ \sigma_{01} & \sigma_1^2 \end{bmatrix}
\end{pmatrix}.
\end{align*}


```{r}
fit_alcuse_3 <-
  brm(data = alcohol,
      alcuse ~ 0 + Intercept + age_14 + coa + age_14:coa + (1 + age_14 | id),
      prior = c(prior(normal(0, 15), class = b)),
      file = "models/fit_alcuse_3")
```



```{r}
fit_alcuse_3
```

```{r}
mcmc_plot(fit_alcuse_3)
```



```{r}
loo_alcuse_2 <- loo(fit_alcuse_2)
loo_alcuse_3 <- loo(fit_alcuse_3)
```

```{r}
loo_compare(loo_alcuse_2, loo_alcuse_3)
```


```{r}
plot(loo_alcuse_3)
```



# Time-varying predictors

If we have time-varying predictor variable, it is a good idea the decompose these into a subject'specific mean (trait) and a session-by-session deviation from that mean (state).

Let's look at some simulated data, in which patient are assigned to either a control or a therapy (ACT) group. Patients' wellbeing (`score`) is assessed over the course of 12 sessions. In addition, we have the patient-rated `therapeutic alliance`, which reflects how the quality of patient-therapist interactions.


```{r}
# https://raw.githubusercontent.com/awellis/learnmultilevelmodels/main/data/wellbeing.csv
wellbeing <- read_csv("data/wellbeing.csv")
```

```{r}
wellbeing |> 
  ggplot(aes(session, score, color = therapy)) +
  geom_line() +
  geom_point() +
  geom_line(aes(session, alliance), linetype = 2) +
  scale_x_continuous(breaks = seq(2, 12, by = 2)) +
  scale_color_brewer(type = "qual") +
  facet_wrap(~patient)
```
We want to predict patient's wellbeing as a function of the therapeutic alliance, while accounting for a linear trend.

:::exercise
- How would you specify this model, as well as unconditional models?
- At which level is the predictor variable `alliance`?
:::



## Decomposing time-varying predictors


Instead of using the raw `alliance` variable, we will create two new variables: the 
average alliance, and the session-by-session deviations from the average. The average will be a patient-level (level 2) predictor, and can be used to explain variability between patients, whereas the deviations reflect fluctuations within patients at individual sessions.


```{r}
wellbeing <- wellbeing |> 
  group_by(patient) |> 
  mutate(al_between = round(mean(alliance, na.rm = TRUE), 2),
         al_within = alliance - al_between)
```



```{r}
fit_alliance_1 <- brm(score ~ session + (1 + session| patient),
                    data = wellbeing,
                    file = "models/fit_alliance_1")
```


```{r}
fit_alliance_2 <- brm(score ~ session + al_between + al_within + 
                      (1 + session  + al_within | patient),
                    data = wellbeing,
                    file = "models/fit_alliance_2")
```

```{r}
fit_alliance_3 <- brm(score ~ session + al_within + 
                      (1 + session  + al_within | patient),
                    data = wellbeing,
                    file = "models/fit_alliance_3")
```



```{r}
fit_alliance_2
```

