---
title: "Response Time Models"
description: | 
  Models for right-skewed data
date: 06-11-2021
author:
  - first_name: "Andrew"
    last_name: "Ellis"
    url: https://github.com/awellis
    affiliation: Cognitive psychology, perception & methods, Univerity of Bern 
    affiliation_url: https://www.kog.psy.unibe.ch
    orcid_id: 0000-0002-2788-936X

citation_url: https://awellis.github.io/learnmultilevelmodels/response-time-models.html
bibliography: ./bibliography.bib
output: 
    distill::distill_article:
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: false
      css: ./css/style.css
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r packages}
library(kableExtra)
library(viridis)
library(tidyverse)
library(brms)

# set ggplot theme
theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))

# set rstan options
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = 4)
```




:::exercise
Check out this interactive web app for response time distributions [@lindelovReactionTimeDistributions2021].

[Reaction time distributions: an interactive overview](https://lindeloev.shinyapps.io/shiny-rt/)
:::


We will try to fit a distribution that is appropriate for response times to data from @wagenmakersLinearRelationMean2007.

Subjects performed a lexical decision task in two conditions (speed stress / accuracy stress). We will look at conly correct responses.

```{r}
library(rtdists)
data(speed_acc) 

speed_acc <- speed_acc %>%
  as_tibble()


df_speed_acc <- speed_acc %>% 
   # remove rts under 180 ms and over 3 sec
  filter(rt > 0.18, rt < 3) %>% 
   # convert to char
  mutate(across(c(stim_cat, response), as.character)) %>% 
  # Kcorrect responses
  filter(response != 'error', stim_cat == response) %>% 
  # convert back to factor
  mutate(across(c(stim_cat, response), as_factor))
```


```{r}
df_speed_acc
```


Plot a few subjects to get an overview.

```{r}
data_plot <- df_speed_acc %>% 
  filter(id %in% c(1, 8, 11, 15))

data_plot %>% 
  ggplot(aes(x = rt)) + 
    geom_histogram(aes(fill = condition), alpha = 0.5, bins = 60) + 
    facet_wrap(~id) +
    coord_cartesian(xlim=c(0, 1.6)) +
    scale_fill_viridis(discrete = TRUE, option = "E")
```



## Shifted Lognormal

We'll not try to fit a shifted-lognormal distribution


The density is given by:

$$ f(x) = \frac{1} {(x - \theta)\sigma\sqrt{2\pi}} exp\left[ -\frac{1}{2} \left(\frac{ln(x - \theta) - \mu }{\sigma}  \right)^2  \right] $$ 


m$\mu$, $\sigma$ und $\theta$ are the three parameters. $\theta$ shifts the entire distribition along the x axis. $\mu$ and  $\sigma$ have an effect on the shape.  



$\mu$ is often interpreted as a **difficulty** parameter. Both median and mean depend on $\mu$

The median is given by $\theta + exp(\mu)$.     


$\sigma$ is the **scale** parameter (standard deviation); It affects the mean, but not the median.


The mean is $\theta + exp\left( \mu + \frac{1}{2}\sigma^2\right)$


We are mainly interested in the $\mu$ parameter. 

When fitting with brms, we need to be aware that $\mu$ and $\sigma$ are on the log scale.



$$ y_i \sim Shifted Lognormal(\mu, \sigma, \theta) $$

$$ log(\mu) = b_0 + b_1 X_1 + ...$$
Both  $\sigma$ and $\theta$ may also be predicted, or simply estimated.
 

$$  \sigma \sim Dist(...) $$

$$  \theta \sim Dist(...) $$



```{r}
priors <- get_prior(rt ~ condition + (1|id), 
          family = shifted_lognormal(),
          data = df_speed_acc)

priors %>% 
  as_tibble() %>% 
  select(1:4)
```


```{r}
priors <- prior(normal(0, 0.1), class = b)
```


```{r}
m1 <- brm(rt ~ condition + (1|id),
          family = shifted_lognormal(),
          prior = priors,
          data = df_speed_acc,
          file = 'models/m1_shiftedlognorm')
```


```{r}
summary(m1)
```




```{r}
# intial values for expected ndt
inits <- function() {
  list(Intercept_ndt = -1.7)
}

m2 <- brm(bf(rt ~ condition + (1|id),
              ndt ~ 1 + (1|id)),
          family = shifted_lognormal(),
          prior = priors,
          data = df_speed_acc,
          inits = inits,
          init_r = 0.05,
          control = list(max_treedepth = 12),
          file = 'models/m2_shiftedlognorm_ndt')
```


```{r}
pp_check(m2, 
         type = "dens_overlay_grouped", 
         group = "condition",
         nsamples = 200) 
```


