---
title: "Intro to Bayesian Statistics"
session: 1
subtitle: "Part 1 <br/> Gentle introduction"
author: "Andrew Ellis"
# institute: "Kognitive Psychologie, Wahrnehmung und Methodenlehre, Universität Bern"
institute: "Bayesian multilevel modelling workshop 2021"
date: 05-21-2021
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["css/xaringan-themer.css", "css/slides-style.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:10
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r child = "setup.Rmd"}
```

```{r load-packages, include=FALSE, warning=FALSE}
library(tidyverse)
library(rmarkdown)
library(countdown)
library(kableExtra)

theme_set(theme_grey(base_size = 14) +
            theme(panel.grid = element_blank()))
```

## Frequentist statistics

Relies on:

- point estimation
- summary statistics
- often uses null hypothesis significance testing

Problems:

- p-values can be hard to understand
- confidence intervals are hard to understand
- it is unclear whether p-values and confidence intervals really allow us to address the questions we care about.
    + what is the probability that my hypothesis might be true?
    + how can I quantify evidence against a null hypothesis?


---

## Example: t-test


We want to compare two groups. One group is wearing fancy hats, the other a the control group. We are interested in their creativity scores.

.panelset[
.panel[.panel-name[Parameters]

```{r}
library(tidyverse)
library(kableExtra)

set.seed(12)
# Number of people per group
N <- 50 
# Population mean of creativity for people wearing fancy hats
mu_fancyhats <- 103 
# Population mean of creativity for people wearing no fancy hats
mu_nofancyhats <- 98 
# Average population standard deviation of both groups
sigma <- 15 
```

]

.panel[.panel-name[Make dataframe]

```{r}
# Generate data
fancyhats = tibble(Creativity = rnorm(N, mu_fancyhats, sigma),
               Group = "Fancy Hat")
nofancyhats = tibble(Creativity = rnorm(N, mu_nofancyhats, sigma),
                 Group = "No Fancy Hat")
FancyHat <- bind_rows(fancyhats, nofancyhats)  %>%
    mutate(Group = fct_relevel(as.factor(Group), "No Fancy Hat"))
```
]

.panel[.panel-name[Data]

```{r}
FancyHat
```
]
.panel[.panel-name[Plot data]

```{r echo=FALSE}
FancyHat %>% 
    ggplot() +
    geom_boxplot ((aes(y = Creativity, x = Group))) +
    labs(title= "Box Plot of Creativity Values")
```

]
]


---

.panelset[
.panel[.panel-name[Welch test]

```{r}
fancyhat_ttest <- t.test(Creativity ~ Group,
       var.equal = FALSE,
       data = FancyHat)
```

]
.panel[.panel-name[Results]

```{r}
fancyhat_ttest_tab <- broom::tidy(fancyhat_ttest)
```

```{r}
fancyhat_ttest_tab %>%
    select(estimate, estimate1, estimate2, statistic, p.value, conf.low, conf.high) %>%
    round(3) %>% 
    kbl() %>%
    kable_classic(full_width = FALSE, html_font = "Cambria")
```

]
.panel[.panel-name[What have we done here?]

1) We estimated two means (and two standard deviations). More specifically, we obtained **point estimates**. 

2) We estimated the difference in means (again, a point estimate).

3) We computed a test statistic..

4) We computed the probability of obtaining a value for the test statistic that is at least as extreme as the one obtained. This is called a p-value.

]]

---

.your-turn[

- Can you explain what the p-value and confidence interval mean? 
- What can you conclude from this analysis?
- Can you think of any problems that might be associated with this type of approach?
]

```{r echo=FALSE}
countdown(minutes = 3)
```

---


## Interpretations of Probability

- In the classical, frequentist approach, parameter, e.g. the means estimated above, do not have probability distributions.
- Only events that can be repeated infinitely many times have a probability, and probability is simply relative frequency.
- In the Bayesian worldview, probability quantifies *degree of belief*. More specificcally, our uncertainty is expressed as a probability distribution. Probability quantifies knowledge, and is not a fundamental property of things.

---

## Some Gamma distributions

```{r gamma-dist, echo=FALSE, layout="l-body-outset"}
library(tidyverse)

tibble(x = seq(from = 0, to = 30, by = .1)) %>% 
  tidyr::expand(x, nesting(alpha = c(2, 3), 
                    beta  = c(0.9, 0.4))) %>% 
  mutate(density = dgamma(x, alpha, beta),
         group   = rep(letters[1:2], times = n() / 2)) %>% 
  
  ggplot(aes(x = x, ymin = 0, ymax = density, 
             group = group, fill = group)) +
  geom_ribbon(size = 0, alpha = 3/4) +
  scale_fill_viridis_d(option = "B", direction = -1, 
                       begin = 1/3, end = 2/3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_cartesian(xlim = c(0, 30)) +
    xlab("") +
  theme(panel.grid = element_blank(),
        legend.position = "none")
```

---


## Bayesian inference

- Parameters have probability distributions.
- Parameters have prior distributions. These quantify our belief before we see the data.
- We obtain posterior distributions instead of point estimates. Posterior distributions 
reflect our belief after having observed data.
- We go from prior to posterior by applying Bayes theorem.
- **Most important point**: Uses probability to quantify uncertainty.


---

## Why should you use Bayesian inference?

- More fun
- Much easier to understand
- Corresponds to our intuitive understanding
- More principled apporach
- Generally much more flexible
- Better for fitting complex models (including multilevel models)
- Allows us to quantify evidence


---

## Why shouldn't you use Bayesian inference?

- Everyone else is using frequentist methods
- Frequentist methods have fancy names for everything (i.e. established off-the-shelf methods)
- Bayesian inference is computationally expensive (as you will soon discover)
- Hypothesis testing is difficult (but the same applies to NHST)
